---
title: "R Practice"
author: "AEL"
date: "6/2/2022"
output: html_document
---

##Intro to Markdown 

Markdown allows you to **bold**, _italicize_, and more!

Make a list with asterisks like:
\*bold with double asterisks 
\*italics with underscores
\*code-type with back ticks

You can also use hyphens instead of asterisks

For numbered lists:
\. bold...
\. italics
\. code...

Hyperlinks are [Hyperlinked Text](https://mylink.org) 

```{r load_data}
library(gapminder)
gap <- gapminder 
```


```{r make-plot}
plot(lifeExp ~ year, data=gap)
```

```{r}
temp <- c(1,2,3)
```

```{r make_cats}
cats <- data.frame(coat = c("calico","black","tabby"),weight = c(2.1,5.0,3.2),likes_string = c(1,0,1))
write.csv(x = cats, file = "original data/feline-data.csv", row.names=FALSE) #saved in folder "original data"
```

```{r read_cats}
read.csv('original data/feline-data.csv',stringsAsFactors = T)
```

Creating vectors and coercing types 

```{r}
my_vector <- vector(mode = 'character',length = 3)
combine_vector <- c(1,2,"bananas")
```

Change names

```{r}
names(cats)[3] <- "likes_hunting"
```

adding  column and rows
```{r}
cats <- read.csv('original data/feline-data.csv')
age <- c(2,3,5)
cats <- cbind(cats,age)
newRow <- list("tortoiseshell",3.3,TRUE,9)
cats <- rbind(cats,newRow)
```

```{r}
human_age <- cats$age*7
as.numeric(human_age)
human_age/7
#by multiplying cats$age by 7 it gives us the cats age in human years, as.numeric converts it to numbers rather than characters, and dividing human_age by 7 gives us the cats original age.
```

Remove rows and columns

```{r}
cats$coat <- as.factor(cats$coat)
cats <- rbind(cats,c("white",3.4,0,1))
cats <- cats[-5,]
cats <- na.omit(cats)
cats <- cats[,-4]
drop <- names(cats) %in% c("weight")
cats[,!drop]
#to remove a column, place a negative before the x value. to remove a row, place a negative infront of the y value
```